---
title: "Comparison of normalisation methods for Human whole muscle and mitochondrial miRNA "
author: "Megan Soria"
date: 2023-07-24
output:
  html_document:
    theme: lumen
    toc: yes
    toc_float: true
  html_notebook:
    theme: lumen
    toc: yes
    toc_float: true
  prettydoc::html_pretty:
    theme: hpstr
    highlight: github
---

This notebook aims to compare normalisation methods for micro-RNAs (miRNAs) extracted from whole skeletal muscle and mitrochodrial fractions from 12 participants of an excercise experiment. Each sample was taken pre, post, and 3hours post exercise. 

```{r include=FALSE}
# load libraries
library(RUVSeq)
library(edgeR)
library(DESeq2)
library(tidyverse)
library(RColorBrewer)
library(RNAseqQC)
library(ggplot2)
library(VennDiagram)
```

# Load counts data and filter out zero counts and 
```{r}
# read in column datafile aka sample information file
muscle.coldata <- read.csv("data/silver_muscle condition.csv", sep = ",", header = TRUE) # do not incl row.names=1
# change "-" into "." to match with the counts labels
muscle.coldata <- muscle.coldata %>% arrange(., desc(condition)) %>%
  mutate(sampleid = gsub("-", ".", sampleid)) %>%
  mutate(sampleid = gsub(" ", "", sampleid))

# muscle counts
muscle.cts <- read.csv("data/silver_muscle raw.csv", sep=",", row.names = 1) #keeps miR identifiers as rownames
muscle.cts[is.na(muscle.cts)] <-0 #replace NA with 0 #now a num not int
muscle.cts <- muscle.cts[, muscle.coldata$sampleid]
muscle.cts <- as.matrix(muscle.cts)
# filter low reads by requiring more than 5 reads in at least two samples for each gene
filter1 <- apply(muscle.cts, 1, function(x) length(x[x>5])>=2)
muscle.cts_f <- muscle.cts[filter1,]

# add factors
muscle.coldata$condition <- factor(muscle.coldata$condition)
muscle.coldata$sampleid <- factor(muscle.coldata$sampleid)
# not filtered coldata
muscle.coldata.nf <- muscle.coldata
muscle.coldata <- muscle.coldata[muscle.coldata$sampleid %in% colnames(muscle.cts_f),]
rownames(muscle.coldata) <- muscle.coldata$sampleid

all(muscle.coldata$sampleid == colnames(muscle.cts_f)) #check these both come back as TRUE

# mito
mito.cts <- read.csv("data/silver_hmt raw_by group.csv", sep=",", row.names = 1) #keeps miR identifiers as rownames
mito.cts[is.na(mito.cts)] <-0 #replace NA with 0 #now a num not int

# remove sample outliers
# filter low reads by requiring more than 5 reads in at least two samples for each gene
filter2 <- apply(mito.cts, 1, function(x) length(x[x>5])>=2)
mito.cts_f <- mito.cts %>%
  select(-c("hmtb.12.1_S29_L001.td", "hmta.12.2.2_S30_L001.tdf", "hmtb.12.3_S31_L001.tdf", "hmtb.11.2_S27_L001.tdf", "hmtb.11.3_S28_L001.tdf")) %>%
  .[filter2,]
mito.cts_f <- as.matrix(mito.cts_f)

#read in column datafile aka sample information file
mito.coldata <- read.csv("data/silver_hmt condition_by group.csv", sep = ",", header = TRUE) # do not incl row.names=1
# change "-" into "." to match with the counts labels
mito.coldata <- mito.coldata %>%
  mutate(sampleid = gsub("-", ".", sampleid)) %>%
  mutate(sampleid = gsub(" ", "", sampleid))
mito.coldata$condition <- factor(mito.coldata$condition)
mito.coldata$sampleid <- factor(mito.coldata$sampleid)
rownames(mito.coldata) <- mito.coldata$sampleid
# not filtered coldata
mito.coldata.nf <- mito.coldata

mito.coldata <- mito.coldata[mito.coldata$sampleid %in% colnames(mito.cts_f),]
all(mito.coldata$sampleid == colnames(mito.cts_f)) #check these both come back as TRUE

datalist = list(mito.coldata.nf = mito.coldata.nf, 
                mito.cts=mito.cts, 
                muscle.coldata.nf=muscle.coldata.nf, 
                muscle.cts=muscle.cts)
saveRDS(datalist, "rds/miRNAdatasets.RDS")
rm(datalist)

miRNAdatasets_trimmed = list(mito.coldata = mito.coldata, 
                mito.cts_f=mito.cts_f, 
                muscle.coldata=muscle.coldata, 
                muscle.cts_f=muscle.cts_f)
saveRDS(miRNAdatasets_trimmed, "rds/miRNAdatasets_trimmed.RDS")
```

# Compare normalisation methods using Relative log expression (RLE) plots (Gandolfo and Speed, 2018) 

## Muscle
```{r fig.width=8, fig.height=5}
# normalise via TMM (edgeR)
d <- DGEList(counts = miRNAdatasets_trimmed$muscle.cts_f, group=muscle.coldata[,-1])
d <- calcNormFactors(d, method="TMM") # get tmm from edgeR
d = estimateCommonDisp(d, verbose=TRUE)
tmm_muscle <- d$pseudo.counts

# VSD from DESEq
dds <- DESeqDataSetFromMatrix(countData = miRNAdatasets_trimmed$muscle.cts_f,
                              colData = miRNAdatasets_trimmed$muscle.coldata,
                              design = ~ condition)
muscle.vsd <- varianceStabilizingTransformation(dds, blind=F) 

set1 <- newSeqExpressionSet(as.matrix(miRNAdatasets_trimmed$muscle.cts_f),
                            phenoData = AnnotatedDataFrame(data = 
                                        miRNAdatasets_trimmed$muscle.coldata, 
                            varMetadata = data.frame(labelDescription=c("sampleid", 
                                                                        "condition"))))
# upper quantile normalisation
set_upper.muscle <- betweenLaneNormalization(set1, which="upper")

set_tmm <- newSeqExpressionSet(as.matrix(tmm_muscle),
                            phenoData = AnnotatedDataFrame(data
                                      = miRNAdatasets_trimmed$muscle.coldata, 
                            varMetadata = data.frame(labelDescription=c("sampleid", "condition"))))

set_vsd <- newSeqExpressionSet(as.matrix(round(assay(muscle.vsd))),
                            phenoData = AnnotatedDataFrame(data = 
                                        miRNAdatasets_trimmed$muscle.coldata, 
                            varMetadata = data.frame(labelDescription=c("sampleid", 
                                                                        "condition"))))

x <- c(rep(1,11), rep(2,12), rep(3,12))
colors <- brewer.pal(3, "Set1") 

plotRLE(set1, outline=FALSE, ylim=c(-3, 3), col=colors[x], xlab="No Normalisation", xaxt = "n")

plotRLE(set_upper.muscle, outline=FALSE, ylim=c(-3, 3), col=colors[x], xlab="Upper qunatile", xaxt = "n") 

plotRLE(set_tmm, outline=FALSE, ylim=c(-3, 3), col=colors[x], xlab="TMM", xaxt = "n")

plotRLE(set_vsd, outline=FALSE, ylim=c(-1, 1), col=colors[x], xlab="VST", xaxt = "n")

plotPCA(tmm_muscle, col=rep(1:3, times = c(11, 12, 12)), labels=FALSE)

pdf(file = "figures/miRNA_musc_normalisation.pdf",   
    width = 8, # The width of the plot in inches
    height = 5) # The height of the plot in inches
plotRLE(set1, outline=FALSE, ylim=c(-3, 3), col=colors[x], xlab="No Normalisation", xaxt = "n")
plotRLE(set_upper.muscle, outline=FALSE, ylim=c(-3, 3), col=colors[x], xlab="Upper qunatile", xaxt = "n") 
plotRLE(set_tmm, outline=FALSE, ylim=c(-3, 3), col=colors[x], xlab="TMM", xaxt = "n")
plotRLE(set_vsd, outline=FALSE, ylim=c(-1, 1), col=colors[x], xlab="VST", xaxt = "n")
plotPCA(tmm_muscle, col=rep(1:3, times = c(11, 12, 12)), labels=FALSE)
dev.off()
```
Based on the RLE plots grouped by condition, the best performing normalisation method for this dataset is TMM, followed by Upper quantile, then VST.

## Mitochondria
```{r fig.width=8, fig.height=5}
# normalise via TMM (edgeR)
d <- DGEList(counts = miRNAdatasets_trimmed$mito.cts_f, group=mito.coldata[,-1])
d <- calcNormFactors(d, method="TMM") # get tmm from edgeR
d = estimateCommonDisp(d, verbose=TRUE)
tmm_mito <- d$pseudo.counts

# VSD from DESEq
dds <- DESeqDataSetFromMatrix(countData = miRNAdatasets_trimmed$mito.cts_f,
                              colData = miRNAdatasets_trimmed$mito.coldata,
                              design = ~ condition)
mito.vsd <- varianceStabilizingTransformation(dds, blind=F) 

set1 <- newSeqExpressionSet(as.matrix(miRNAdatasets_trimmed$mito.cts_f),
                            phenoData = AnnotatedDataFrame(data=
                                      miRNAdatasets_trimmed$mito.coldata, 
                            varMetadata = data.frame(labelDescription=c("sampleid", 
                                                                        "condition"))))
# upper quantile normalisation
set_upper.mito <- betweenLaneNormalization(set1, which="upper")

set_tmm <- newSeqExpressionSet(as.matrix(tmm_mito),
                            phenoData = AnnotatedDataFrame(data=
                                      miRNAdatasets_trimmed$mito.coldata, 
                            varMetadata = data.frame(labelDescription=c("sampleid", 
                                                                        "condition"))))

set_vsd <- newSeqExpressionSet(as.matrix(round(assay(mito.vsd))),
                            phenoData = AnnotatedDataFrame(data=
                                        miRNAdatasets_trimmed$mito.coldata, 
                            varMetadata = data.frame(labelDescription=c("sampleid", 
                                                                        "condition"))))

x <- c(rep(1,8), rep(2,10), rep(3,10))
colors <- brewer.pal(3, "Set1")

plotRLE(set1, outline=FALSE, ylim=c(-5, 5), col=colors[x], xlab="No Normalisation", xaxt = "n")

plotRLE(set_upper.mito, outline=FALSE, ylim=c(-3.5, 3.5), col=colors[x], xlab="Upper qunatile Normalisation", xaxt = "n") 

plotRLE(set_tmm, outline=FALSE, ylim=c(-3, 3), col=colors[x], xlab="Trimmed mean of M values (TMM)", xaxt = "n")

plotRLE(set_vsd, outline=FALSE, ylim=c(-1, 1), col=colors[x], xlab="Variance Stabilising Transformation", xaxt = "n")

plotPCA(tmm_mito, col=rep(1:3, times = c(9, 12, 12)), labels=FALSE)

pdf(file = "figures/miRNA_mito_normalisation.pdf",   
    width = 15, # The width of the plot in inches
    height = 12) # The height of the plot in inches
plotRLE(set1, outline=FALSE, ylim=c(-5, 5), col=colors[x], xlab="No Normalisation", xaxt = "n")
plotRLE(set_upper.mito, outline=FALSE, ylim=c(-3.5, 3.5), col=colors[x], xlab="Upper qunatile Normalisation", xaxt = "n") 
plotRLE(set_tmm, outline=FALSE, ylim=c(-3, 3), col=colors[x], xlab="Trimmed mean of M values (TMM)", xaxt = "n")
plotRLE(set_vsd, outline=FALSE, ylim=c(-1, 1), col=colors[x], xlab="Variance Stabilising Transformation", xaxt = "n")
plotPCA(tmm_mito, col=rep(1:3, times = c(9, 12, 12)), labels=FALSE)
dev.off()

```
Based on the RLE plots grouped by condition, the best performing normalisation method for this dataset is Upper quantile, followed by TMM, then VST.


# Density plot for gene counts of mitochondria vs muscle miRNAs
```{r fig.width=8, fig.height=4}
# muscle
select2 <- as.matrix(assay(muscle.vsd))

#pull out samples by group or time-point
PRE <- select2[,1:11]
POST_0h <- select2[,12:23]
POST_3h <- select2[,24:35]

#row (i.e. miRNA) means by group or time-point
rmeansPRE <- as.vector(rowMeans(PRE))
rmeansPOST_0h <- as.vector(rowMeans(POST_0h))
rmeansPOST_3h <- as.vector(rowMeans(POST_3h))

df.muscle <- data.frame(rmeansPRE, rmeansPOST_0h, rmeansPOST_3h) %>% 
  mutate(Sample = "Muscle") %>% 
  mutate(genes = row.names(select2))


# mitochondria
select2 <- as.matrix(assay(mito.vsd)) 

#pull out samples by group or time-point
PRE <- select2[,1:8]
POST_0h <- select2[,9:18]
POST_3h <- select2[,19:28]

#row (i.e. miRNA) means by group or time-point
rmeansPRE <- as.vector(rowMeans(PRE))
rmeansPOST_0h <- as.vector(rowMeans(POST_0h))
rmeansPOST_3h <- as.vector(rowMeans(POST_3h))

df.mito <- data.frame(rmeansPRE, rmeansPOST_0h, rmeansPOST_3h) %>% 
  mutate(Sample = "Mito") %>% 
  mutate(genes = row.names(select2))

p <- function(v) {
  if(length(v) == 2){
    f=paste(v[1], "&", v[2])
  }else(f=paste(v))
  return(f)
 }
# pre muscle and mito
PRE.combined <- df.mito %>% 
  select(rmeansPRE, Sample, genes) %>%
  bind_rows(select(df.muscle, rmeansPRE, Sample, genes)) %>%
  group_by(genes) %>%
  summarise(rmeansPRE = mean(rmeansPRE),
            Sample = p(as.character(Sample)))

# 3hr post muscle and mito
POST3h.combined <- df.mito %>% 
  select(rmeansPOST_3h, Sample, genes) %>%
  bind_rows(select(df.muscle, rmeansPOST_3h, Sample, genes)) %>%
  group_by(genes) %>%
  summarise(rmeansPOST_3h = mean(rmeansPOST_3h),
            Sample = p(as.character(Sample)))

# post muscle and mito
POST.combined <- df.mito %>% 
  select(rmeansPOST_0h, Sample, genes) %>%
  bind_rows(select(df.muscle, rmeansPOST_0h, Sample, genes)) %>%
  group_by(genes) %>%
  summarise(rmeansPOST_0h = mean(rmeansPOST_0h),
            Sample = p(as.character(Sample)))

# plot
pdf(file = "figures/miRNAs_densityPlots.pdf",   
    width = 8, # The width of the plot in inches
    height = 4) # The height of the plot in inches

theme_set(theme_classic(base_size = 12))
ggplot(data=PRE.combined, aes(x=rmeansPRE, group=Sample, fill=Sample)) +
    geom_density(adjust=1.5, alpha=.75) + 
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Normalised counts of miRNAs in humans:",
       subtitle = "mitochondira and muscle pre-exercise") +
    theme(plot.title=element_text(hjust=0.5),
          plot.subtitle=element_text(hjust=0.5)) 
  
ggplot(data=POST.combined, aes(x=rmeansPOST_0h, group=Sample, fill=Sample)) +
    geom_density(adjust=1.5, alpha=.75) +
    scale_fill_brewer(palette = "Set1") +
    labs(title = "Normalised counts of miRNAs in humans:",
       subtitle = "mitochondira and muscle post-exercise") +
     theme(plot.title=element_text(hjust=0.5),
          plot.subtitle=element_text(hjust=0.5))

ggplot(data=POST3h.combined, aes(x=rmeansPOST_3h, group=Sample, fill=Sample)) +
    geom_density(adjust=1.5, alpha=.75) +
    scale_fill_brewer(palette = "Set1") +
    labs(title = "Normalised counts of miRNAs in humans:",
       subtitle = "mitochondira and muscle 3 hours post-exercise") +
     theme(plot.title=element_text(hjust=0.5),
          plot.subtitle=element_text(hjust=0.5))

dev.off()
```

# Dataset preparation for differential expression analysis
Use TMM normalisation and DESeq pipeline.
Objects from the data loading chunk (Chunk 2) will be re-used.
The original, non-filtered muscle counts data and coldata will be saved in the final rds object. 

```{r}
# filter out the outlier sample but not the low counts; this will be done later after creating the DESeq object

# mito
mito.cts_f <- mito.cts %>%
  select(-c("hmtb.12.1_S29_L001.td", "hmta.12.2.2_S30_L001.tdf", "hmtb.12.3_S31_L001.tdf", "hmtb.11.2_S27_L001.tdf", "hmtb.11.3_S28_L001.tdf"))
mito.cts_f <- as.matrix(mito.cts_f)

# use the non-filtered mito coldata - mito.coldata.nf - as input
mito.coldata_f <- mito.coldata.nf[mito.coldata.nf$sampleid %in% colnames(mito.cts_f),]
all(mito.coldata$sampleid == colnames(mito.cts_f)) #check these both come back as TRUE

# save the datasets for DE analysis
miRNAdatasets_forDE = list(mito.coldata_f = mito.coldata_f, 
                mito.cts_f=mito.cts_f, 
                muscle.coldata=muscle.coldata, 
                muscle.cts=muscle.cts)
saveRDS(miRNAdatasets_forDE, "rds/miRNAdatasets_forDE.RDS")
rm(miRNAdatasets_forDE)
```

Rerferences:
Gandolfo, L.C. and Speed, T.P. (2018) ‘RLE plots: Visualizing unwanted variation in high dimensional data’, PLoS ONE, 13(2), p. e0191629. Available at: https://doi.org/10.1371/journal.pone.0191629.

Session Info
```{r session info}
sessionInfo()
```

