---
title: 'Skeletal Muscle DE Analysis'
author: "Jessica Silver (additional analysis by Megan Soria)"
date: 2023-06-16
output:
  html_document:
    theme: lumen
    toc: yes
    toc_float: true
  html_notebook:
    theme: lumen
    toc: yes
    toc_float: true
  prettydoc::html_pretty:
    theme: hpstr
    highlight: github
---

source: https://github.com/jessicasilver/miRNAs

## Summary and background information
The data used in this analysis is from an exercise intervention study in 12 human participants conducted by **[name]** on **[year]**. Muscle biopsies were taken at 3 time points for each participant: pre-exercise, post-exercise, and 3 hours post-exercise <br> <br>
A whole muscle and a mitochondrial fraction is extracted from each biopsy. This analysis focuses on the differential expression of the whole muscle transcriptome.  

## Load and install packages
The following packages are required for the following analysis and visualisation. The code below will automatically install and load packages from CRAN. Packages from Bioconductor can be installed by un-commenting (removing the # before the line or selecting the code block and pressing ctrl+shift+c) and running  the code block for the Bioconductor packages. Sometimes these installations will throw an error and specific packages might need to be installed manually.

```{r include=FALSE}
suppressPackageStartupMessages({c(
  # packages from CRAN
  if(!require("prettydoc")) install.packages("prettydoc"), library(prettydoc),
  if(!require("readxl")) install.packages("readxl"), library(readxl),
  if(!require("dplyr")) install.packages("dplyr"), library(dplyr),
  if(!require("RColorBrewer")) install.packages("RColorBrewer"), library(RColorBrewer),
  if(!require("circlize")) install.packages("circlize"), library(circlize),
  library(ggplot2),
  library(RNAseqQC),
  # packages from Bioconductor
  # run the installation code when needed
  # if (!require("BiocManager", quietly = TRUE)) install.packages("BiocManager")
  #   BiocManager::install(c("DESeq2", "ComplexHeatmap", "EnhancedVolcano", "org.Hs.eg.db")),
  library(DESeq2),
  library(ComplexHeatmap),
  library(EnhancedVolcano),
  library(cowplot),
  library(PCAtools)
  )})

# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("DESeq2")
  
# load self-authored functions 
source("functions.R")
```

***

## Differential expression analysis
The following differential expression analysis is conducted using the DESeq2 package

```{r DEseq2,echo=T, results='hide', message=FALSE}
data <- read.csv("data/2022-11-11_human skm_total RNA_RAW.csv", sep=",", row.names = 1) #alternatively, use raw rno-gastroc totalRNA seq file

data[is.na(data)] <-0 #replace NA with 0  
cts <- as.matrix(data) #convert to matrix

#read in column datafile aka sample information file
coldata <- read.csv("data/silver_hskm condition_by group.csv", sep = ",", header = TRUE) 

coldata <- coldata[,c("sampleid","condition")]
coldata$condition <- factor(coldata$condition)
coldata$sampleid <- factor(coldata$sampleid)

all(rownames(coldata$sampleid) %in% colnames(cts)) 
all(rownames(coldata$sampleid) == colnames(cts)) 

#set up deseq2 object
dds <- DESeqDataSetFromMatrix(countData = cts,
                              colData = coldata,
                              design = ~ condition)

dds$condition <- relevel(dds$condition, ref = "PRE") #set your baseline. For rno-gastroc, this will be `ref = "SEDENTARY"`

keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]

dds <- DESeq(dds) #run deseq2

#extract the results for each comparison that you are interested in 
res <- results(dds, contrast=c("condition","3HPOST","PRE")) 

resOrdered <- res[order(res$pvalue),]
summary(resOrdered)
head(resOrdered)

```

#### Pull out DEseq2 normalised read counts

```{r deseq2_norm, results='hide'}
dds <- estimateSizeFactors(dds)
normcounts <- counts(dds, normalized=TRUE)
```

### Extract all possible contrasts 

```{r}
# resNames_list() function sourced from function.R
ddsResults <- resNames_list(dds)

# extract 3HRPOST vs POST contrast
hr3POST_vs_POST <- results(dds, contrast=c("condition","3HPOST","POST")) %>%
  .[order(.$log2FoldChange, decreasing = TRUE),] %>%
      as.data.frame(.)
condition_3HPOST_vs_POST <- list("condition_3HPOST_vs_POST" = hr3POST_vs_POST)

# store all contrasts in a list 
ddsResults_list <- append(ddsResults, condition_3HPOST_vs_POST)
ddsResults_list <- diff_expressed(ddsResults_list, padjust = 0.05, L2FC = 1)
names(ddsResults_list)

saveRDS(ddsResults_list, "rds/ddsResults_wholeMusc.rds")

# threshFilter() function sourced from function.R
ddsSignif_contrasts <- threshFilter(ddsResults_list, padjust = 0.05, L2FC = 1)
saveRDS(ddsSignif_contrasts, "rds/ddsSignif_wholeMusc.rds")

```

## Visualisations

Prepare data and add gene names
```{r }
# add gene names from ENSEMBL Ids
data_genename <- read.csv("data/2022-11-11_human skm_total RNA_RAW_ENSEMBL-geneID.csv", sep=",", row.names = 1)
data_genename <- data_genename[,1:2] 

signif_genes <- left_join(ddsSignif_contrasts$condition_3HPOST_vs_PRE %>%
                            mutate(ENSEMBL = rownames(ddsSignif_contrasts$condition_3HPOST_vs_PRE)),
                          data_genename, by = 'ENSEMBL') 
                                  
rownames(signif_genes) <- signif_genes$ENSEMBL
signif_genes <- signif_genes %>% dplyr::select(-c(ENSEMBL))
```

Apply variance stabilising transformation to the DESeq data
```{r fig.width=10, fig.height=6}
# apply variance stabilising transformation to the DESeq data
# this normalisation is required for calculations involving distances (heatmaps, clustering, etc.)
vsd <- vst(dds, blind=F)

#plotPCA(vsd, intgroup=c("condition"))
pca <- plot_pca(vsd, PC_x = 1, PC_y = 2, color_by = "condition", point_rel_size = 4) 
p <- pca(assay(vsd))

s <- screeplot(p, axisLabSize = 18, titleLabSize = 22) 
s

pdf(file = "figures/wholeMusc_mRNA_pca.pdf",
    width = 10, # The width of the plot in inches
    height = 6) # The height of the plot in inches
pca
s
dev.off()

```

### Heatmap
This heatmap shows the changes in gene expression across all time points using condition = POST vs PRE. The top 30 up-regulated and down-regulated genes are filtered according to pvalue < 0.05 and |log2 fold change| > 1
```{r}
mat <- assay(vsd)[rownames(signif_genes),]

# get he z scores for the matrix
mat_z <- t(apply(mat, 1, scale))
colnames(mat_z) <- colnames(mat)

head(mat_z)

# select the top and bottom genes and store log2foldchange and baseMean values
numkeep <- 70
rowskeep <- c(seq(1:numkeep))

l2_val <- as.matrix(signif_genes[rowskeep,]$log2FoldChange)
colnames(l2_val) <- "logFC"

bMean <- as.matrix(signif_genes[rowskeep,]$baseMean)
colnames(bMean) <- "AveExpr"

# colours for values between b/w/r for min/max of l2 values
col_logFC <- colorRamp2(c(min(l2_val), 0, max(l2_val)), c("blue", "white", "red"))

# colours for 0% quantile, and 75% quantile of mean values --- 0, 25, 50, 75, 100
col_AveExpr <- colorRamp2(c(quantile(bMean)[1], quantile(bMean)[4]), c("white", "red"))

# set a pre, post, and 3hour post groupings
sampling_group <- rep(c("Pre", "Post", "3hrPost"), times = c(10, 12, 11))
sampling_group_col = c("Pre" = 5, "Post" = 4, "3hrPost" = 6)

dend1 = cluster_within_group(mat_z, sampling_group)
```

Create the heatmaps
```{r fig.width=14.22, fig.height=15}
ha <- HeatmapAnnotation(summary = anno_summary(gp = gpar(fill = 2),
                                               height = unit(2, "cm")))

h1_noGrouping <- Heatmap(mat_z[rowskeep,], cluster_rows = F, # cluster only columns to compare samples
            column_labels = colnames(mat_z), name="Z-score",
            cluster_columns = T)

h1 <- Heatmap(mat_z[rowskeep,], cluster_rows = F, # cluster only columns to compare samples
            column_labels = colnames(mat_z), row_labels = signif_genes$SYMBOL[rowskeep], name="Z-score",
            cluster_columns = dend1, column_split = 3,
            top_annotation = HeatmapAnnotation(Sampling = sampling_group, col = list(Sampling = sampling_group_col)))

h2 <- Heatmap(l2_val, row_labels = signif_genes$SYMBOL[rowskeep], 
            cluster_rows = F, name="logFC", col = col_logFC,
            #cell_fun = function(j, i, x, y, w, h, col) { # add text to each grid
              #grid.text(round(l2_val[i, j],2), x, y)}
            )

h3 <- Heatmap(bMean, row_labels = signif_genes$SYMBOL[rowskeep], 
             cluster_rows = F, name = "AveExpr", col=col_AveExpr)

hA<-h1+h2+h3
hA

hB <- h1_noGrouping+h2+h3
hB

pdf(file = "figures/wholeMusc_mRNA_hmap.pdf",   
    width = 14.22, # The width of the plot in inches
    height = 15) # The height of the plot in inches
hA
hB
dev.off()
```
### Volcano plots
```{r fig.width=14.22, fig.height=15}
plots <- list()
for (i in 1:length(ddsResults_list)) {
   
  upGenes <- table(ddsResults_list[[i]]$diffexpressed)["UP"]
  downGenes <- table(ddsResults_list[[i]]$diffexpressed)["DOWN"]
  name1 <- gsub("_", " ", names(ddsResults_list[i]))
  name <- gsub("condition", "", name1)
    
    p <- EnhancedVolcano(ddsResults_list[[i]], x="log2FoldChange", y="padj", lab = NA, 
                pCutoff = 0.05, FCcutoff = 1, title = name, pointSize = 4,
                subtitle = paste0("Up-regulated = ", upGenes, " and Down-regulated = ", downGenes), 
                gridlines.major = FALSE, gridlines.minor = FALSE, axisLabSize = 20, titleLabSize = 20, subtitleLabSize = 18,
                captionLabSize = 0, legendLabSize = 15, legendIconSize = 8, col = c("grey30", "royalblue", "forestgreen", "red2"))
  
  plots[[i]] <- p
}
plot_grid(plotlist = plots, ncol = 2, labels = LETTERS[1:length(plots)], label_size = 18)

pdf(file = "figures/wholeMusc_mRNA_volcanoPlots.pdf",   
    width = 10, # The width of the plot in inches
    height = 10) # The height of the plot in inches
plots[1:3]
dev.off()
```

### Summary of up-regulated and down-regulated genes
```{r }
diffexpr_summary <- list()

for (i in 1:length(ddsResults_list)) {
  
  upGenes <- table(ddsResults_list[[i]]$diffexpressed)["UP"]
  downGenes <- table(ddsResults_list[[i]]$diffexpressed)["DOWN"]
  name <- gsub("_", " ", names(ddsResults_list[i]))
  name2 <- gsub("condition", "", name)
  
  diffexpr_summary[[paste0(i)]]$Stage <- name2
  diffexpr_summary[[paste0(i)]]$Up <- upGenes
  diffexpr_summary[[paste0(i)]]$Down <- downGenes
}

diffexpr_summaryDF <- do.call(rbind.data.frame, diffexpr_summary) %>%
  as.data.frame(.)
write.csv(diffexpr_summaryDF, "outData/wholeMusc_mRNA_DEsummary.csv")
write.csv(ddsResults_list$condition_3HPOST_vs_PRE, "outData/wholeMusc_mRNA_DE_3HPOSTvPRE.csv")
write.csv(ddsResults_list$condition_POST_vs_PRE, "outData/wholeMusc_mRNA_DE_POSTvPRE.csv")
write.csv(ddsResults_list$condition_3HPOST_vs_POST, "outData/wholeMusc_mRNA_DE_3HPOSTvPOST.csv")
```

Session Info
```{r session info} 
sessionInfo()
```

