---
title: "Rat Mitochondria Target Prediction and Over-Representation Analysis"
author: "Jessica Silver"
date: "2022-11-18"
output: html_document
theme: paper
  
---

source: https://github.com/jessicasilver/miRNAs

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Set Up Instructions for DE analysis and visualisation 

Prior to running, ensure that `multiMiR` and `readxl` packages are installed. 

```{r setupinstructions}
suppressPackageStartupMessages({
library(multiMiR)
library(readxl) #only needed if using .xlsx files but not .csv
})

```

***

#### Target Prediction analysis using the multiMiR package

For target prediction analysis, a list of candidate miRNAs was first provided. Provide full miRNA identifier in the mature form, i.e. `hsa-miR-1-3p`. Note that analysis was also replicated for `table = "predicted", predicted.site = "all",` to identify predicted miRNA targets with conserved and
nonserved regions in the 3' UTR of target mRNAs.

```{r candidate miRNAs}
mircandidates <- read_excel("2022-11-20_rno-miR candidates.xlsx", 
                            sheet = 1, 
                            na = c("","NA", "?", "#VALUE!", "Undetermined"),
                            trim_ws = TRUE)  
Vector <- c(mircandidates$miR) #multimiR requires a vector to run

expval_rnocand <- get_multimir(org = "rno", 
                          mirna = Vector, 
                          table = "validated", 
                          summary = T)


expval_rnocand_dupl <- expval_rnocand@data #pull out data with duplicates
expval__rnocand_genelist <- unique(expval_rnocand_dupl) #remove duplicates
write.csv(expval__rnocand_genelist, file = "2022-11-20_rno_candidate-genelist_expval_targets.csv")
```

Next, the "univerise" list of genes was provided and included all miRNAs detected above thresholds in isolated mitochondria.

```{r background miRNAs}
mir_ALL <- read_excel("2022-11-20_rno_full list.xlsx", 
                      sheet = 1, 
                      na = c("","NA", "?", "#VALUE!", "Undetermined"),
                      trim_ws = TRUE)  

Vector_all <- c(mir_ALL$miR) #multimiR requires a vector to run

expval_rnoALL <- get_multimir(org = "rno", 
                               mirna = Vector_all, 
                               table = "validated", 
                               summary = T)

expval_rnoALL_dupl <- expval_rnoALL@data #with duplicates
expval__rnoALL_genelist <- unique(expval_rnoALL_dupl)
write.csv(expval__rnoALL_genelist, file = "2022-11-20_rno_ALL-genelist_expval_targets.csv")
```

***

#### ORA 

Prior to running, ensure that `clusterProfiler` and `enrichplot` packages are installed. Install the relevant organism data base `org.Rn.eg.db` for 
rat mito ORA.  

```{r setupinstructions}
suppressPackageStartupMessages({
library(clusterProfiler)
library(enrichplot) 
library("org.Rn.eg.db")
})

```


Next, over representation analysis (ORA) using `clusterProfiler v4.4.4` was used for gene target pathway enrichment to predict which biological 
functions may be affected by the candidate miRNAs. 

Initially, the background gene list used for the ORA analysis comprised the intersect of all gene symbols identified as “experimentally validated” 
or “predicted” targets of candidate miRNAs, and all genes detected in skeletal muscle from the participants in the current study

```{r skeletal muscle transcripts}
rno_skmtranscripts <- read.csv("rno-gastroc_totalRNA_rawfiltered_geneSYMBOL.csv", sep=",") #keeps miR identifiers as rownames
rno_skmSYMBOL <- rno_skmtranscripts$SYMBOL #this is your list of transcripts in skel muscle in the current study

#find common elements between experimentally validated gene list `target_symbol`
#`gene`
expval_rnocand_geneID <- expval__rnocand_genelist$target_symbol
expval_rnocand_skmspecific_geneID <- Reduce(intersect, list(rno_skmSYMBOL,expval_rnocand_geneID)) 
#write.csv! 

#`universe`
expval_rnoALL_geneID <-expval__rnoALL_genelist$target_symbol
expval_rnoALL_skmspecific_geneID <- Reduce(intersect, list(rno_skmSYMBOL,expval_rnoALL_geneID))
#write.csv! 
```

Then, run your ORA

```{r experimentally validated ORA in skel muscle}
GO_rno_expval_skmspecific <- enrichGO(gene         = expval_rnocand_skmspecific_geneID,
                          OrgDb         = org.Rn.eg.db, 
                          keyType       = 'SYMBOL',
                          ont           = "all",
                          pAdjustMethod = "BH",
                          universe = expval_rnoALL_skmspecific_geneID, #Background Genes
                          qvalueCutoff  = 0.05)

GO_rno_expval_skmspecific_res <- GO_rno_expval_skmspecific@result
write.csv(GO_rno_expval_skmspecific_res, file = "2022-11-20_rno-miR_gene enrichment_expval-skeletalmuscle.csv")
```

Next, a secondary ORA analysis included a background gene list comprised of all gene symbols identified as “experimentally validated” or “predicted”
targets of candidate miRNAs, without specifically selecting those detected in skeletal muscle tissue in the current study. The tissueEnrich package was
used to show the expression profiles of tissue-specific genes in the secondary ORA analysis.

```{r experimentally validated ORA non skeletal muscle specific}
expval_rnocand_ENSEMBL <- expval__rnocand_genelist$target_ensembl
expval_rnoALL_ENSEMBL <- expval__rnoALL_genelist$target_ensembl

GO_rno_expval_global <- enrichGO(gene         = expval_rnocand_ENSEMBL,
                                 OrgDb         = org.Rn.eg.db, #change to human
                                 keyType       = 'ENSEMBL',
                                 ont           = "all",
                                 pAdjustMethod = "BH",
                                 universe = expval_rnoALL_ENSEMBL, #Background Genes
                                 qvalueCutoff  = 0.05)
#export
GO_rno_expval_global_res <- GO_rno_expval_global@result
write.csv(GO_rno_expval_global_res, file = "2022-11-20_rno-miR_gene enrichment_expval-global.csv")
```

Below are full scripts following the above steps (global and skeletal muscle-specific) for predicted miRNA targets

```{r predicted ALL scripts}
predALL_rnocand <- get_multimir(org = "rno", 
                               mirna = Vector, #`genes`
                               table = "predicted", predicted.site = "ALL", 
                               summary = T)

predALL_rnocand_dupl <- predALL_rnocand@data #with duplicates
predALL__rnocand_genelist <- unique(predALL_rnocand_dupl) #remove duplicates
write.csv(predALL__rnocand_genelist, file = "2022-11-20_rno_candidate-genelist_predALL_targets.csv")

predALL_rnoALL <- get_multimir(org = "rno", 
                              mirna = Vector_all, #`universe`
                              table = "predicted", predicted.site = "ALL", 
                              summary = T)
predALL_rnoALL_dupl <- predALL_rnoALL@data #with duplicates
predALL__rnoALL_genelist <- unique(predALL_rnoALL_dupl) #remove duplicates
write.csv(expval__rnoALL_genelist, file = "2022-11-20_rno_ALL-genelist_predALL_targets.csv")

#ORA - global (not specific for skeletal muscle genes)
predALL_rnocand_ENSEMBL <- predALL__rnocand_genelist$target_ensembl
predALL_rnoALL_ENSEMBL <- predALL__rnoALL_genelist$target_ensembl

GO_rno_predALL_global <- enrichGO(gene         = predALL_rnocand_ENSEMBL,
                                 OrgDb         = org.Rn.eg.db, 
                                 keyType       = 'ENSEMBL',
                                 ont           = "all",
                                 pAdjustMethod = "BH",
                                 #universe = predALL_rnoALL_ENSEMBL, #Background Genes
                                 qvalueCutoff  = 0.05)
GO_rno_predALL_global_res <- GO_rno_predALL_global@result #get results
write.csv(GO_rno_predALL_global_res, file = "2022-11-20_rno-miR_gene enrichment_predALL-global_nouniverse.csv") #export

#ORA - skeletal muscle specific
#find common elements between predALL gene list `target_symbol`
#`gene`
predALL_rnocand_geneID <- predALL__rnocand_genelist$target_symbol
predALL_rnocand_skmspecific_geneID <- Reduce(intersect, list(rno_skmSYMBOL,predALL_rnocand_geneID)) 
#write.csv! 

#`universe`
predALL_rnoALL_geneID <-predALL__rnoALL_genelist$target_symbol
predALL_rnoALL_skmspecific_geneID <- Reduce(intersect, list(rno_skmSYMBOL,predALL_rnoALL_geneID))
#write.csv!

GO_rno_predALL_skmspecific <- enrichGO(gene         = predALL_rnocand_skmspecific_geneID,
                                      OrgDb         = org.Rn.eg.db, 
                                      keyType       = 'SYMBOL',
                                      ont           = "all",
                                      pAdjustMethod = "BH",
                                      universe = predALL_rnoALL_skmspecific_geneID, #Background Genes
                                      qvalueCutoff  = 0.05)
GO_rno_predALL_skmspecific_res <- GO_rno_predALL_skmspecific@result #get results
write.csv(GO_rno_expval_skmspecific_res, file = "2022-11-20_rno-miR_gene enrichment_predALL-skeletalmuscle.csv") #export
```


#### Visualise
There a number of different ways to visualise the ORA outputs. Cover all bases by ensuring the following packages are installed. 

```{r setupinstructions visualisation}
suppressPackageStartupMessages({
library(ggnewscale)
library(tidyverse)
library(TissueEnrich)
library(tidyr)
})

```

**Option 1:** simple dot plots. 

```{r ORA dotplots}
dotplot(GO_rno_expval_global)
dotplot(GO_rno_expval_skmspecific)
dotplot(GO_rno_predALL_global)
dotplot(GO_rno_predALL_skmspecific)
```

**Option 2:** tree plots. Example below for "experimentally validated" miRNA targets found in secondary (non-skeletal muscle specific) analysis. 

```{r ORA treeplot}
#1. experimentally validated - global
edox_expval_global <- pairwise_termsim(GO_rno_expval_global)
treeplot(edox_expval_global, 
         #label_format = 3, 
         nCluster =5,
         nWords =5, #Number of words for Cluster Tags 
         offset= rel(2), #Moves the dendogram clustering
         offset_tiplab =rel(2), # Moves GO pathways
         label_format_cladelab=30,
         extend=0.3,
         hexpand =0.3, #Moves position of cluster tags
         #group_color = c("#FF0000", "#000099", "#336600", "#990099", "#990033"),
         fontsize=4,
         geneClusterPanel = "heatMap")+
  ggplot2::theme(legend.position=c(0.95,#across  
                                   0.5), #Up/down 
                 text = element_text(size = 10)) 

#2. experimentally validated - skeletal muscle-specific gene list
edox_expval_skel <- pairwise_termsim(GO_rno_expval_skmspecific)
treeplot(edox_expval_skel, 
         #label_format = 3, 
         nCluster =5,
         nWords =5, #Number of words for Cluster Tags 
         offset= rel(2), #Moves the dendogram clustering
         offset_tiplab =rel(2), # Moves GO pathways
         label_format_cladelab=30,
         extend=0.3,
         hexpand =0.3, #Moves position of cluster tags
         #group_color = c("#FF0000", "#000099", "#336600", "#990099", "#990033"),
         fontsize=4,
         geneClusterPanel = "heatMap")+
  ggplot2::theme(legend.position=c(0.95,#across  
                                   0.5), #Up/down 
                 text = element_text(size = 10)) 

#3. predicted - global
edox_predALL_global <- pairwise_termsim(GO_rno_predALL_global)
treeplot(edox_predALL_global, 
         #label_format = 3, 
         nCluster =5,
         nWords =5, #Number of words for Cluster Tags 
         offset= rel(2), #Moves the dendogram clustering
         offset_tiplab =rel(2), # Moves GO pathways
         label_format_cladelab=30,
         extend=0.3,
         hexpand =0.3, #Moves position of cluster tags
         #group_color = c("#FF0000", "#000099", "#336600", "#990099", "#990033"),
         fontsize=4,
         geneClusterPanel = "heatMap")+
  ggplot2::theme(legend.position=c(0.95,#across  
                                   0.5), #Up/down 
                 text = element_text(size = 10)) 

#4. predicted - skeletal muscle specific 
treeplot(edox_predALL_skel, 
         #label_format = 3, 
         nCluster =2,
         nWords =0, #Number of words for Cluster Tags 
         offset= rel(2), #Moves the dendogram clustering
         offset_tiplab =rel(2), # Moves GO pathways
         #label_format_cladelab=30,
         extend=0.3,
         hexpand =0.3, #Moves position of cluster tags
         #group_color = c("#FF0000", "#000099", "#336600", "#990099", "#990033"),
         fontsize=4,
         geneClusterPanel = "heatMap")+
  ggplot2::theme(legend.position=c(0.77,#across  
                                   0.5), #Up/down 
                 text = element_text(size = 10))

```

**Other Notes: ** You can pull out the geneID (ENSEMBL) for a specific pathway using the code chunk below. Note that this does not exported in other
results tables. 
Note that the example given below is a code chunk from analysis conducted on human mito small RNA seq data. This was not performed on the rat 
mito samples. 

```{r multiMiR and ORA extras}
geneID <- GO_hsa_expval2@result$geneID
descr <- GO_hsa_expval2@result$Description

test <- data.frame(descr,geneID)
write.csv(test, file = "yyyy-mm-dd_hsa-miR_ennrichGO_expval2_geneIDdescr.csv")
```

You can search by individual miRNAs. See code chunk below. All other downstream processing steps (removing duplicates, running `enrichGo` and
visualisation) are identical to those detailed for candidate miRNAs above. 
Same as above chunk - this example using human data but the process is the same for rat. 

```{r multiMiR by individual miRNA identifier}
expval_146b3p = get_multimir(org = "hsa", #"rno"
                             mirna = "hsa-miR-146b-3p", #"rno-miR-1-3p" #etc
                             #(can also use miR Base accession number or both)
                             table = "validated", 
                             summary = T)
```

Session Info
```{r session info}
sessionInfo()
```
